Almost compatible functions and infinite length games

### 2015-12-03 articles/almost-compatible-functions-and-topological-games/article.tex
paper complete modulo editing of Dow's results and topological implications

  \begin{definition}
    The statement \(\alcompSP{\kappa}\) weakens \(\alcompS{\kappa}\) be only requiring
    the witness functions \(f_A:A\to\omega\) to be finite-to-one.
  \end{definition}

@@

  \begin{definition}
    A Kurepa family \(\mc K\subseteq[\kappa]^{\aleph_0}\) on \(\kappa\)
    satisfies that
    \(\mc K\rest A=\{K\cap A:K\in\mc K\}\) is countable
    for each \(A\in[\kappa]^{\aleph_0}\).
  \end{definition}

  \begin{theorem}
    \(\alcompSP{\kappa}\) holds whenever there exists a cofinal Kurepa family on \(\kappa\).
  \end{theorem}

@@

  We now construct a topology on \(\omega_n\) for each \(n<\omega\) which
  will witness a Kurepa family of size \(\aleph_n\). A similar construction
  was previously shown in \cite{MR539228}.

@@

  \begin{definition}
    A topological space is said to be \(\omega\)-bounded if each countable
    subset of the space has compact closure.
  \end{definition}

  We note that in
  \cite{MR539228} a \(T_2\) locally countable \(\omega\)-bounded space
  was called \term{splendid}, and the claim that there exists a splendid
  space of cardinality \(\kappa\) was called \(S(\kappa)\).

@@

  \begin{theorem}
    For each \(k< \omega\), there is a \(T_2\), locally countable,
    \(\omega\)-bounded topology on \(\omega_k\). (That is, \(S(\omega_k)\)
    for all \(k<\omega\)).
  \end{theorem}

@@

  \begin{lemma}
    The family of compact open sets in a locally countable, \(\omega\)-bounded
    topological space is a Kurepa family cofinal in \([X]^\omega\).
  \end{lemma}

@@

  The above lemma may alternatively be stated as ``\(S(\kappa)\) implies
      the existence of a cofinal Kurepa family on \(\kappa\)''.

  \begin{corollary}
    There exists a Kurepa family cofinal in \([\omega_k]^\omega\)
    for each \(k<\omega\).
  \end{corollary}

  This corollary is also an observation of Todorcevic
  communicated by Dow in \cite{MR1229125}:
  if every Kurepa family of size at most \(\theta\)
  extends to a cofinal Kurepa family, then the same is true of \(\theta^+\).
  So the result follows as
  every Kurepa family \(\mc K\) of size \(\omega\) extends to
  the cofinal Kurepa family \([\bigcup\mc K]^\omega\).

  With this we have our desired result.

  \begin{corollary}
    \(\alcompSP{\aleph_k}\) holds for all \(k<\omega\). Under \(CH\), we have
    both \(\alcompSP{\aleph_2}\) and \(\neg\alcompS{\aleph_2}\).
  \end{corollary}

  As noted in [TODO cite Dow],
  Jensen's one-gap two-cardinal theorem under \(V=L\) [TODO cite] can be used
  to show that there exist cofinal Kurepa families on every cardinal.

  \begin{corollary}[\(V=L\)]
    \(\alcompSP{\theta}\) holds for all cardinals.
  \end{corollary}

  In particular, \(\alcompS{\omega_2}\) fails under \(CH\), showing the two are
  distinct. Actually, \(CH\) is not required to have \(\alcompS{\omega_2}\) fail.

  % TODO: read through Alan's proof and fit to rest of document
        We are going to need a technical lemma (available in Kunen).
        \bigskip

        \begin{lemma}
        Assume that $G\subset \operatorname{Fn}(\omega_2,2)$ is a generic filter,
         and let $\mu\in \omega_2$. Then the final model $V[G]$ can be
         regarded as forcing  with $\operatorname{Fn}(\omega_2\setminus \mu,
         2)$ over the model $V[G_\mu]$.
        In addition, for each $\operatorname{Fn}(\omega_2,2)$-name $\dot A$
        of a subset of $\omega$ (treat as a subset of $\omega\times
        \operatorname{Fn}(\omega_2,2) $),
        there is a canonical name $\dot A(G_\mu)$ where,
        $$\dot A(G_\mu) = \{ (n,p\restriction [\mu,\omega_2))  :
         (n,p)\in \dot A\ \ \mbox{and} \ \ p\restriction \mu\in G_\mu\}$$
        and we get that the valuation of $\dot A(G_\mu)$ by the tail
        of the generic, $G_{\omega_2\setminus \mu}$, is the same as
        the valuation of $\dot A$ by the full generic.
        \end{lemma}

        \begin{theorem}
          If we add $\omega_2$ Cohen reals to a model of CH we get
        that Scheepers' $S(\omega_2)$ (still) fails.
        \end{theorem}

@@

  On the other hand, it's also consistent that \(\alcompSP{\theta}\) can fail.

  \begin{theorem}
    There's a model where \(\alcompSP{\omega_\omega}\)
    fails.
  \end{theorem}

@@

          We will need the following claim from \cite{MR1045371}

@@

             Since $\mathbb P_0$ has cardinality less than $|N|=\kappa^{+\omega+1}$,
              the sequence $\{ h_\xi : \xi \in \Lambda\cap N\}$ remains unbounded
              mod finite in $\Pi_n \varrho_n$ (and in
               $\Pi_n (\varrho_n\cap N)$).  Now pass to the extension by
                $G\cap \mathbb P_0$ and let $H$ be the function
                 $\operatorname{val}_{G}(\dot H)$, and we recall
                 that $f_{\zeta_\xi}(h_\xi(n)) = H(n)$ for all $n\in \omega$.
           Now pass to the full extension $V[G]$ and again, since
            $\mathbb P_1$ was forced to be countably closed,
             the family $\{ h_\xi : \xi \in \Lambda\cap N\}$ is still
             unbounded in $\Pi_n (\varrho_n \cap N)$. We let
             $A$ be the countable set $N\cap \lambda^{+\omega}$,
             and for  each
             $\xi\in \Lambda\cap N$, there is an $n_\xi$ such that
              $f_{\xi}(h_\xi(m)) = f_A(h_\xi(m))$ for all $m>n_\xi$.
              There is a single $n$ so that $\Lambda_n
               = \{\xi\in \Lambda\cap N       : n_\xi = n\}$ has cardinality
                $\omega_1 $, and thus
                $\{ h_\xi : \xi\in \Lambda_n\cap N\}$ is also unbounded
                in $\Pi_n (\rho_n\cap N)$.
          This certainly implies that there is an $m>n$
          such that $\{ h_\xi (m) : \xi\in \Lambda_n\cap N\}$ is infinite.
          This completes the proof since  $f_{A}(h_\xi(m)) = H(m)$
          for all $\xi\in \Lambda_n\cap N$.

  \end{proof}

  \begin{question}
    Is \(\alcompSP{\theta}\) equivalent to having a Kurepa family on \(\theta\)?
  \end{question}

>>>>

  \begin{definition}
    The statement \(\alcompSP{\kappa}\) weakens \(\alcompS{\kappa}\) by only
    requiring the witnessing almost-compatible functions \(f_A:A\to\omega\)
    to be finite-to-one.
  \end{definition}

@@

  \begin{definition}
    A Kurepa family \(\mc K\subseteq[\kappa]^{\aleph_0}\) on \(\kappa\)
    satisfies that
    \(\mc K\rest A=\{K\cap A:K\in\mc K\}\) is countable
    for each \(A\in[\kappa]^{\aleph_0}\).
    Let \(\kurepaS\kappa\) be the statement claiming there exists
    a Kurepa family on \(\kappa\)
    cofinal in \([\kappa]^{\aleph_0}\).
  \end{definition}

  \begin{theorem}
    \(\kurepaS\kappa \Rightarrow \alcompSP{\kappa}\).
  \end{theorem}

@@

  We now construct a topology on \(\omega_n\) for each \(n<\omega\) which
  will witness a Kurepa family of size \(\aleph_n\). A similar construction
  was previously shown by Juh\'asz et. al. in \cite{MR539228},
  and the relationship of
  Kurepa families and such spaces has also been investigated in a preprint
  of Nyikos \cite{NYIKOSKUREPA}.

@@

  \begin{definition}
    A topological space is said to be \(\omega\)-bounded if each countable
    subset of the space has compact closure.
    As in \cite{MR539228} we call a
    \(T_2\), locally countable, \(\omega\)-bounded space
    \term{splendid}, and let \(\splendidS\kappa\) represent
    the claim that there exists a splendid space of cardinality \(\kappa\).
  \end{definition}

@@

  \begin{theorem}
    For each \(k< \omega\), there is a \(T_2\), locally countable,
    \(\omega\)-bounded topology on \(\omega_k\). That is, \(\splendidS{\aleph_k}\)
    for all \(k<\omega\).
  \end{theorem}

@@

  \begin{lemma}
    The family of compact open sets in a locally countable, \(\omega\)-bounded
    topological space \(X\)
    is a Kurepa family cofinal in \([X]^\omega\). That is,
    \(\splendidS\kappa \Rightarrow \kurepaS\kappa\).
  \end{lemma}

@@

  \begin{corollary}
    \(\kurepaS{\aleph_k}\) for all \(k<\omega\).
  \end{corollary}

  Alternatively, the previous corollary may be obtained
  via an observation of Todorcevic
  communicated by Dow in \cite{MR1229125}:
  if every Kurepa family of size at most \(\kappa\)
  extends to a cofinal Kurepa family, then the same is true of \(\kappa^+\).

  Nyikos points out in \cite{NYIKOSKUREPA} that a cofinal Kurepa family may
  be used to construct a locally metrizable, \(\omega\)-bounded,
  zero-dimensional space with appropriate cardinality,
  but whether this can be
  strengthened to locally countable and \(\omega\)-bounded (as asked in
  \cite{MR539228}) remains an open question.

  Also left open is this extension of the question asked in \cite{NYIKOSKUREPA}
  and \cite{MR539228} on the possible equivalence of \(\splendidS\kappa\)
  and \(\kurepaS\kappa\).

  \begin{question}
    May any of the implications in the theorem
    \(\splendidS\kappa\Rightarrow\kurepaS\kappa\Rightarrow\alcompSP\kappa\)
    be reversed?
  \end{question}

  Regardless, we have obtained our desired result.

  \begin{corollary}
    \(\alcompSP{\aleph_k}\) for all \(k<\omega\).
  \end{corollary}

  \section{More consistency results}

  As noted in \cite{MR1229125}, Jensen's one-gap two-cardinal theorem under
  \(V=L\) introduced in \cite{MR0309729} implies that \(\kurepaS\kappa\)
  holds for all cardinals \(\kappa\).

  \begin{corollary}[\(V=L\)]
    \(\alcompSP{\kappa}\) for all cardinals \(\kappa\).
  \end{corollary}

  Weakening to the continuum hypothesis, we have an obvious consequence.

  \begin{corollary}[\(CH\)]
    \(\alcompSP{\aleph_2}\), but \(\neg\alcompS{\aleph_2}\).
  \end{corollary}

  But \(CH\) is not required to have \(\alcompS{\aleph_2}\) fail.

  % TODO: read through Alan's proofs and fit to rest of document

            \begin{lemma}[\cite{MR597342}]
            Assume that $G\subset \operatorname{Fn}(\omega_2,2)$ is a generic filter,
             and let $\mu\in \omega_2$. Then the final model $V[G]$ can be
             regarded as forcing  with $\operatorname{Fn}(\omega_2\setminus \mu,
             2)$ over the model $V[G_\mu]$.
            In addition, for each $\operatorname{Fn}(\omega_2,2)$-name $\dot A$
            of a subset of $\omega$ (treat as a subset of $\omega\times
            \operatorname{Fn}(\omega_2,2) $),
            there is a canonical name $\dot A(G_\mu)$ where,
            $$\dot A(G_\mu) = \{ (n,p\restriction [\mu,\omega_2))  :
             (n,p)\in \dot A\ \ \mbox{and} \ \ p\restriction \mu\in G_\mu\}$$
            and we get that the valuation of $\dot A(G_\mu)$ by the tail
            of the generic, $G_{\omega_2\setminus \mu}$, is the same as
            the valuation of $\dot A$ by the full generic.
            \end{lemma}

            \begin{theorem}
              There exists a model of \(ZFC\) for which \(\mf c=\aleph_2\)
              and \(\neg\alcompS{\aleph_2}\).
            \end{theorem}

@@

  We are also able to force \(\alcompSP{\kappa}\) to fail for some cardinal.

  \begin{theorem}
    There exists a model of \(ZFC\) for which \(\neg\alcompSP{\aleph_\omega}\).
  \end{theorem}

@@

          We will need the following claim from \cite{MR1045371}:

@@

             Since $\mathbb P_0$ has cardinality less than $|N|=\kappa^{+\omega+1}$,
              the sequence $\{ h_\xi : \xi \in \Lambda\cap N\}$ remains unbounded
              mod finite in $\Pi_n \varrho_n$ (and in
               $\Pi_n (\varrho_n\cap N)$).  Now pass to the extension by
                $G\cap \mathbb P_0$ and let $H$ be the function
                 $\operatorname{val}_{G}(\dot H)$, and we recall
                 that $f_{\zeta_\xi}(h_\xi(n)) = H(n)$ for all $n\in \omega$.
           Now pass to the full extension $V[G]$ and again, since
            $\mathbb P_1$ was forced to be countably closed,
             the family $\{ h_\xi : \xi \in \Lambda\cap N\}$ is still
             unbounded in $\Pi_n (\varrho_n \cap N)$. We let
             $A$ be the countable set $N\cap \lambda^{+\omega}$,
             and for  each
             $\xi\in \Lambda\cap N$, there is an $n_\xi$ such that
              $f_{\xi}(h_\xi(m)) = f_A(h_\xi(m))$ for all $m>n_\xi$.
              There is a single $n$ so that $\Lambda_n
               = \{\xi\in \Lambda\cap N       : n_\xi = n\}$ has cardinality
                $\omega_1 $, and thus
                $\{ h_\xi : \xi\in \Lambda_n\cap N\}$ is also unbounded
                in $\Pi_n (\rho_n\cap N)$.
          This certainly implies that there is an $m>n$
          such that $\{ h_\xi (m) : \xi\in \Lambda_n\cap N\}$ is infinite.
          This completes the proof since  $f_{A}(h_\xi(m)) = H(m)$
          for all $\xi\in \Lambda_n\cap N$.
  \end{proof}
